Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2024-03-21 manuscript/manuscript_file.Rmd
response to technical reviews

## Psuedo abstract

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but recent evidence of N~2~O uptake across a broad range of waterbodies have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r national_stats$post_m_prop_sat * 100`% (95% credible interval: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) of waterbodies functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate (NO~3~) concentration, waterbody size, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

Lakes, reservoirs, and ponds (waterbodies hereafter) are considered globally significant sources of nitrous oxide (N~2~O)[@DelSontro_etal_2018a], a potent greenhouse gas and contributor to stratospheric ozone destruction. It is generally assumed that N~2~O emissions from surface waters scale with inorganic nitrogen loading [@Mosier_etal_1998] or productivity [@DelSontro_etal_2018], but recent reports of N~2~O uptake across waterbodies in Canada [@Soued_etal_2016; @Webb_etal_2019] and Finland [@Kortelainen_etal_2020] have called this assumption into question. Furthermore, detailed investigations of individual waterbodies have shown that lentic N~2~O biogeochemistry is dynamic, with some waterbodies alternating between N~2~O sources and sinks in response to nitrogen availability, thermal stratification, and other factors [@Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014].

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Nitrous oxide dynamics have been investigated in streams and rivers spanning a broad range of environmental conditions, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochemistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that were at least 1 hectare in area, at least 1 meter deep, had at least 0.1 hectare of open water, and had a minimum residence time of one week[@EPAdesign] **(Fig. 1)**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r round(((dg %>% dplyr::filter(n2o.src.snk == "sink") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 1)` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **(Fig. 1)**. These data, along with the survey design information, were used to develop a Bayesian hierarchical model to predict N~2~O concentrations for all `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` CONUS waterbodies in the target population. At the population level, the majority of CONUS waterbodies (posterior mean = `r national_stats$post_m_prop_sat * 100`%, 95% CI: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`) were predicted to be a net sink for N~2~O. The Western Mountains ecoregion had the smallest proportion of waterbodies predicted as N~2~O sinks (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(UCL_prop_sat) * 100`) and the Northern Plains had the greatest (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(UCL_prop_sat) * 100`), although the 95% credible intervals over the predictive distributions overlap across all ecoregions **(Table 1)**.

![Figure 1. Location of sampling sites. Color of the points indicates whether the waterbody was functioning as a source or sink of nitrous oxide (N~2~O). Point size indicates the dissolved N~2~O concentration.](manuscript_figures/figure1.png)

The estimated extent of N~2~O undersaturation in CONUS waterbodies was surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long-standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as N~2~O sinks. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months. An investigation of boreal systems in Canada found that 40% of rivers, lakes, and ponds sampled during the summer were functioning as N~2~O sinks[@Soued_etal_2016]. Similarly, 29% of surveyed boreal lakes in Finland were found to be N~2~O sinks during the summer [@Kortelainen_etal_2020]. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern held across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use.

@@

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "total waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% waterbodies)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(metric ton N2O y-1)") %>% # 1 megagram (Mg) = 1 1000 kg = 1 metric ton
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean saturation ratio, undersaturation extent, emission rate, and flux estimates for N2O at the national and ecoregional scales.", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption

  footnote(i = 1, j = 3:6,
           ref_symbols = "a",
           value = as_paragraph("Mean and 95% credible intervals based on 2000 draws from the posterior distribution"),
           part = "header"
  ) %>%
    footnote(i = 1, j = 4,
           ref_symbols = ", b",
           value = as_paragraph("Percent of population with N2O saturation ratio < 1."),
           part = "header",
           inline = TRUE
  ) %>%
  footnote(i = 1, j = 5:6,
           ref_symbols = ", c",
           value = as_paragraph("Negative values indicate N2O uptake"),
           part = "header",
           inline = TRUE
  ) %>%
  fontsize(size = 8, part = "all") %>% # doesn't apply to caption when as_paragraph is invoked.
  fontsize(size = 7, part = "footer") %>% #smaller font for footnote
    autofit() %>% # convenient column spacing
  fit_to_width(ft_1, max_width = 11) # scale to fill 8x11 paper, landscape
```

@@

Despite widespread N~2~O undersaturation across CONUS waterbodies, the estimated means for N~2~O saturation ratio (ratio of measured to equilibrium concentration) were indistinguishable from one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction of widespread N~2~O undersaturation with average dissolved N~2~O concentrations closer to equilibrium can be attributed to the prominent right skew predicted for dissolved N~2~O concentrations across the population of CONUS waterbodies **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, each predicted distribution of N~2~O saturation ratios had a median value less than 1, indicating that most waterbodies were undersaturated in these populations. However, the long right tails on these distributions, comprised of relatively rare waterbodies with very high levels of N~2~O supersaturation, were sufficient to drive the population means to values close to, or even greater than 1. Thus, most waterbodies in the CONUS were functioning as N~2~O sinks during the summer months, but a few functioned as strong N~2~O sources, particularly at high NO~3~ and high water temperature **(Fig. 3)**.

![Figure 2. Predicted distribution of N~2~O saturation values across waterbodies in the nine major ecoregions. Each blue line (n=2000) represents the distribution of values from a single realization of the posterior predictive distribution. Vertical black dashed and solid lines indicate point estimates of the median and mean, respectively, across all realizations (Table 1). Vertical red line indicates a saturation ratio of 1. x-axis is truncated at 3 to better show the area where 97% of the observations occurred.](manuscript_figures/figure2.tiff)

@@

![Figure 3. Estimated interactive effects of surface water temperature (A and B) and waterbody size (B and D) with NO~3~ levels on the N~2~O saturation ratio.  A) The relationship between N~2~O saturation and surface water temperature, dependent upon nitrate (NO~3~) level.  The five, ordered, categorical levels for NO~3~ correspond to concentration ($\mu$g L^-1^) ranges in the observed data of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13. B) The relationship between N~2~O saturation and NO~3~ level conditional on surface water temperature. The NO~3~ effect was estimated at 5 quantiles of the surface temperatures observed (10%, 25%, 50%, 75%, 90%). C) The relationship between N~2~O saturation and waterbody surface area (log-transformed hectares) conditional on NO~3~ level. D) The relationship between N~2~O saturation and NO~3~ level conditional on (log) waterbody surface area (5 quantiles).  Horizontal dashed lines in each panel reflect an N~2~O saturation value of 1.](manuscript_figures/cond_effect_no3.tif)

@@

Nitrous oxide saturation ratio was also related to waterbody size **(Fig. 3)**.  Small waterbodies with low NO~3~ levels were predicted to be undersaturated, whereas small waterbodies with high NO~3~ were predicted to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, were predicted to have N~2~O saturation ratios closer to equilibrium. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreased with increasing waterbody size class **(Fig. 4)**. This pattern holds regardless of whether N~2~O was supersaturated or undersaturated. Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In waterbodies with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column [@Carpenter_1983]. Furthermore, sediment may warm faster in small and shallow waterbodies, whereas thermal stratification may substantially delay sediment warming in large and deep waterbodies, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is supersaturated or undersaturated in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the effect of biogeochemistry on dissolved N~2~O may be more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished. While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018; @Juutinen_etal_2009; @Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

@@

The predicted emission rates for CONUS waterbodies spanned a broad range (95% CI min: `r em_min_max$min_LCL` - `r em_min_max$min_UCL` mg N~2~O m^-2^ d^-1^, 95% CI max: `r em_min_max$max_LCL` - `r em_min_max$max_UCL` mg N~2~O m^-2^ d^-1^), which was expected for this large and diverse collection of waterbodies, but within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **(Fig. 5A)**, which reflects the relationship between N~2~O saturation and waterbody size **(Fig. 3C,D)**. Despite widespread N~2~O undersaturation predicted across CONUS waterbodies, the estimate for the national average N~2~O emission rate was centered over positive values (posterior mean = `r national_stats$post_m_mean_en2o`, 95% CI: `r national_stats$LCL_mean_en2o` - `r national_stats$UCL_mean_en2o` mg N~2~O m^-2^ d^-1^) **(Table 1)** and clearly exceeded the negative estimate for the national median (posterior mean = `r national_stats$post_m_median_en2o`, 95% CI: `r national_stats$LCL_median_en2o` - `r national_stats$UCL_median_en2o` mg N~2~O m^-2^ d^-1^). This departure from the median was due to relatively rare waterbodies with very high emission rates pulling the mean towards the direction of positive skew **(SI Fig. 1)**.

![Figure 5: A) Predicted N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. Points and bars reflect the posterior means (points) and 95% credible intervals, respectively, based on 2000 draws from the posterior predictive distribution for each waterbody. C) A comparison of the estimated relative flux contributions from waterbodies in each size class to the total CONUS flux. The red line connects the mean of the posterior estimate of the ratio, at each size class, based on 2000 draws. The black polygon indicates the 95% credible intervals for the ratio estimates across the size classes](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Our research is the first to suggest that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Our observations of dissolved N~2~O were predicted reasonably well using a regression model with interactions between observed water temperature, NO~3~ concentration, and waterbody size (see SI). These modeling results suggested that NO~3~ levels may play a role in determining whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink may be exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increased with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

>>>>

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere but recent evidence of N~2~O uptake across a broad range of lakes have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r national_stats$post_m_prop_sat * 100`% (95% credible interval: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) of lakes functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate concentration, waterbody size, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS lakes is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS lakes are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

Lakes, reservoirs, and ponds (lakes hereafter) are considered globally significant sources of nitrous oxide (N~2~O)[@DelSontro_etal_2018a], a potent greenhouse gas and contributor to stratospheric ozone destruction. It is generally assumed that N~2~O emissions from surface waters scale with inorganic nitrogen loading [@Mosier_etal_1998] or productivity [@DelSontro_etal_2018], but recent reports of N~2~O uptake across lakes in Canada [@Soued_etal_2016; @Webb_etal_2019], the United States [Aho_et_al_2023], and Finland [@Kortelainen_etal_2020] have called this assumption into question. These recent studies have been conducted at relatively small spatial scales, or with a relatively small number of lakes, and the prevalence of N~2~O uptake in lakes at national to continental scales is unknown.

Nitrous oxide can be produced and consumed via microbially mediated nitrogen transformations occurring in the sediment and water column of lakes.  The balance between N~2~O production and consumption determines whether a lake functions as a N~2~O source or sink to the atmosphere.  Nitrous oxide production and consumption rates are determined by a broad range of factors including inorganic nitrogen, dissolved organic carbon, and oxygen concentrations, among others, making it difficult to predict the balance between these processes[@Beaulieu_etal_2015;@Soued_etal_2016;@Webb_etal_2019]. Furthermore, N~2~O production and consumption rates can vary spatially and temporally within a lake, with some waterbodies even alternating between N~2~O sources and sinks through time[@Beaulieu_etal_2014;@Musenze_etal_2014], adding additional challenges to predicting the source/sink status of lakes.

Here we present results from the first national-scale survey of N~2~O concentrations and emissions from lakes in the continental U.S. (CONUS).  Samples sites were selected using a generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] and the observational data were used to develop a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` CONUS lakes.  Combining the largest freshwater N~2~O dataset with a hierarchical model enabled us to estimate the prevalence of N~2~O uptake in CONUS lakesfor the first time and to identify important environmental controls on N~2~O biogeochemistry.

@@

During the summer of 2017 the United States Environmental Protection Agency conducted the first national-scale survey of N~2~O concentrations and emissions in CONUS lakes as part of the National Lakes Assessment (NLA) [@USEPA_NLA17_2022].  The population of interest (POI) for the survey was `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` CONUS lakes that had a surface area of at least 1 ha.  A representative sample (n=`r dg %>% filter(sample.source == "DG", visit.no == 1, sitetype == "PROB") %>% nrow()`) was selected from the POI using a generalized random-tessellation stratified survey design [@Stevens_Olsen_2004]**(Fig. 1)**. The survey design was stratified by state and used unequal probability categories based on lake size to ensure that large lakes, which are far less common than small lakes, were included in the study.  Sample sites were distributed across all nine major CONUS ecoregions [@Herlihy_etal_2008] and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. The sample data and survey design information (e.g., stratification variables and unequal probability categories) were used to develop a Bayesian hierarchical model to predict dissolved and equilibrium N~2~O concentrations for the POI. Predictor variables for the model were restricted to those required to generate predictions that reproduced the observed N~2~O measures in an out-of-sample dataset and were available for the POI.  See Methods and SI for a detailed discussion of variable selection.

Nearly `r round(((dg %>% dplyr::filter(n2o.src.snk == "sink") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 1)` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **(Fig. 1)**. The Bayesian hierarchical model predicted that the majority lakes in the POI (posterior mean = `r national_stats$post_m_prop_sat * 100`%, 95% credible interval (CI): `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`) functioned as net N~2~O sinks during the summer months. The Western Mountains ecoregion had the smallest proportion of lakes predicted as N~2~O sinks (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(UCL_prop_sat) * 100`) and the Northern Plains had the greatest (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(UCL_prop_sat) * 100`), although the 95% credible intervals over the predictive distributions overlap across all ecoregions **(Table 1)**.

![Figure 1. Location of sampling sites with an underlay of the nine aggregated level III US ecoregions. Color of the points indicates whether the waterbody was functioning as a source or sink of nitrous oxide (N~2~O). Point size indicates the dissolved N~2~O concentration.](manuscript_figures/figure1.png)

The estimated extent of N~2~O undersaturation in CONUS lakes was surprisingly high, particularly for nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long-standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lakes can function as N~2~O sinks. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months. An investigation of boreal systems in Canada found that 40% of rivers, lakes, and ponds sampled during the summer were functioning as N~2~O sinks[@Soued_etal_2016]. Similarly, 29% of surveyed boreal lakes in Finland were found to be N~2~O sinks during the summer [@Kortelainen_etal_2020]. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lakes function as N~2~O sinks during the summer months. This pattern held across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use.

@@

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "total waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% lakes)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(metric ton N2O y-1)") %>% # 1 megagram (Mg) = 1 1000 kg = 1 metric ton
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean saturation ratio, undersaturation extent, emission rate, and flux estimates for N2O at the national and ecoregional scales.", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption

  footnote(i = 1, j = 3:6,
           ref_symbols = "a",
           value = as_paragraph("Mean and 95% credible intervals based on 2000 draws from the posterior distribution"),
           part = "header"
  ) %>%
    footnote(i = 1, j = 4,
           ref_symbols = ", b",
           value = as_paragraph("Percent of population of interest with N2O saturation ratio < 1."),
           part = "header",
           inline = TRUE
  ) %>%
  footnote(i = 1, j = 5:6,
           ref_symbols = ", c",
           value = as_paragraph("Negative values indicate N2O uptake"),
           part = "header",
           inline = TRUE
  ) %>%
  fontsize(size = 8, part = "all") %>% # doesn't apply to caption when as_paragraph is invoked.
  fontsize(size = 7, part = "footer") %>% #smaller font for footnote
    autofit() %>% # convenient column spacing
  fit_to_width(ft_1, max_width = 11) # scale to fill 8x11 paper, landscape
```

@@

Despite widespread N~2~O undersaturation across the POI, estimated means for N~2~O saturation ratio (ratio of measured to equilibrium concentration) were indistinguishable from a value of one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction of widespread N~2~O undersaturation and average dissolved N~2~O concentrations close to equilibrium can be attributed to the prominent right skew in the predicted distribution of dissolved N~2~O concentrations across the POI **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, the predicted distribution of N~2~O saturation ratios had a median value less than 1, indicating that most lakes were undersaturated in these populations. However, the long right tails on these distributions, comprised of relatively rare lakes with very high levels of N~2~O supersaturation, were sufficient to drive the population means to values close to, or even greater than 1. Thus, most lakes in the CONUS were functioning as N~2~O sinks during the summer months, but a few functioned as strong N~2~O sources, particularly at high NO~3~ and high water temperature **(Fig. 3)**.

![Figure 2. Predicted distribution of N~2~O saturation values across lakes in the nine major ecoregions. Each blue line (n=2000) represents the distribution of values from a single realization of the posterior predictive distribution. Vertical black dashed and solid lines indicate point estimates of the median and mean, respectively, across all realizations (Table 1). Vertical red line indicates a saturation ratio of 1. x-axis is truncated at 3 to better show the area where 97% of the observations occurred.](manuscript_figures/figure2.tiff)

@@

![Figure 3. Estimated interactive effects of surface water temperature (A and B) and waterbody size (B and D) with NO~3~ levels on the N~2~O saturation ratio.  A) The relationship between N~2~O saturation and surface water temperature, dependent upon nitrate (NO~3~) level.  The five, ordered, categorical levels for NO~3~ correspond to concentration ($\mu$g N L^-1^) ranges in the observed data of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13. B) The relationship between N~2~O saturation and NO~3~ level conditional on surface water temperature. The NO~3~ effect was estimated at 5 quantiles of the surface temperatures observed (10%, 25%, 50%, 75%, 90%). C) The relationship between N~2~O saturation and waterbody surface area (log-transformed hectares) conditional on NO~3~ level. D) The relationship between N~2~O saturation and NO~3~ level conditional on (log) waterbody surface area (5 quantiles).  Horizontal dashed lines in each panel reflect an N~2~O saturation value of 1.](manuscript_figures/cond_effect_no3.tif)

@@

Lake size was also related to N~2~O saturation ratio, but the relationship was dependent on NO~3~ concentration **(Fig. 3)**.  Small lakes with low NO~3~ levels were predicted to be undersaturated, whereas small lakes with high NO~3~ were predicted to be highly supersaturated. Large lakes, regardless of NO~3~ status, were predicted to have N~2~O saturation ratios closer to equilibrium. . Thus N~2~O saturation ratio tended to deviate from a value of one more strongly in small lakes as compared to larger lakes.  This pattern can also be seen by calculating the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O.  When pooled across all ecoregions, $\Delta$ N~2~O is predicted to decline as lake size increases **(Fig. 4)**. This pattern holds regardless of whether N~2~O was supersaturated or undersaturated. Lake size and shape can affect dissolved gas concentration through several mechanisms. For example, N~2~O inputs from streams or groundwater[@Wang_etal_2023] may constitute a substantial N~2~O subsidy to small lakes but are more likely be diluted in larger lakes with greater volumes. Small lakes also tend to have lower water volume to benthic surface area ratios,and therefore stronger connections between sediment, an area of active N~2~O processing, and the air-water interface where our samples were collected. Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep waterbodies, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller lakes. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated lakes. Regardless of whether a lake is supersaturated or undersaturated in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small lakes, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the effect of biogeochemistry on dissolved N~2~O may be more pronounced in small lakes because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished. While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018; @Juutinen_etal_2009; @Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

@@

The predicted emission rates for the POI spanned a broad range (95% CI min: `r em_min_max$min_LCL` - `r em_min_max$min_UCL` mg N~2~O m^-2^ d^-1^, 95% CI max: `r em_min_max$max_LCL` - `r em_min_max$max_UCL` mg N~2~O m^-2^ d^-1^), which was expected for this large and diverse collection of lakes, but is within the range of values reported in the literature [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing lake size **(Fig. 5A)**, which reflects the relationship between N~2~O saturation and lake size **(Fig. 3C,D)**. Despite widespread N~2~O undersaturation predicted across the POI, the estimate for the national average N~2~O emission rate was centered over positive values (posterior mean = `r national_stats$post_m_mean_en2o`, 95% CI: `r national_stats$LCL_mean_en2o` - `r national_stats$UCL_mean_en2o` mg N~2~O m^-2^ d^-1^) **(Table 1)** and clearly exceeded the negative estimate for the national median (posterior mean = `r national_stats$post_m_median_en2o`, 95% CI: `r national_stats$LCL_median_en2o` - `r national_stats$UCL_median_en2o` mg N~2~O m^-2^ d^-1^). This departure from the median was due to relatively rare lakes with very high emission rates pulling the mean towards the direction of positive skew **(SI Fig. 1)**.

![Figure 5: A) Predicted N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 lakes in the population of interest. Points and bars reflect the posterior means (points) and 95% credible intervals, respectively, based on 2000 draws from the posterior predictive distribution for each waterbody. C) A comparison of the estimated relative flux contributions from lakes in each size class to the total CONUS flux. The red line connects the mean of the posterior estimate of the ratio, at each size class, based on 2000 draws. The black polygon indicates the 95% credible intervals for the ratio estimates across the size classes.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Our research is the first to suggest that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Our observations of dissolved N~2~O were predicted reasonably well using a regression model with interactions between observed water temperature, NO~3~ concentration, and waterbody size (see SI). These modeling results suggested that NO~3~ levels may play a role in determining whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink may be exaggerated with increasing water temperature. These results suggest that the convergence of the ongoing warming of CONUS lakes [@Smucker_etal_2021;@OReilly_etal_2015;@Winslow_etal_2017] and projected increases in freshwater nitrogen loading [@Sinha_etal_2017] may increase lentic N~2~O emissions over the next century[@Wang_etal_2023].

Although lakes in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these lakes is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved and equilibrium N~2~O concentrations translate to large uncertainties in flux when propagated across the surface of large lakes. Uncertainty in the national flux estimate was too large to determine if CONUS lakes functioned as a net source or sink of N~2~O. While our sample included some of the largest lakes in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include an assessment of intra-waterbody spatial variation in N~2~O, which can be important in large waterbodies[@Salk_etal_2019; @Borges_etal_2023], as well as careful accounting of measurement and analytical uncertainty, which can propagate to uncertainties in flux estimates.
